---
title: "GSE74548: Gene Methylation Enrichment Analysis - accounting for correlation"
author: "hcy_meth meta-analysis group"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: yes
    toc_float: true
    code_folding: hide
    fig_width: 7
    fig_height: 7
theme: cosmo
---

##  Introduction

Previously, Mandhri and I analysed the B-PROOF 450K data, trying to understand
whether vitamin supplementation caused changes in gene methylation.
We used Limma and some basic analyses, which showed no specific probes with FDR<0.05,
nor any DMRs.

In this analysis we will use the principle of Gene Set Enrichment Analysis, applying it
to many probes belonging to genes.
If the probes are trending in concert, then we can make some judgement about the
enrichment of those probes.
The statistical test used is the CAMERA test, which is a competitive test that attempts to account for
correlation between genes in a set, or in this case probes belonging to a gene.

```{r,libs}

library("parallel")
library("dplyr")
library("kableExtra")
library("eulerr")
library("mitch")
library("limma")
library("IlluminaHumanMethylation450kmanifest")
library("IlluminaHumanMethylation450kanno.ilmn12.hg19")
library("tictoc")

```

## Get array annotation data

```{r,anno}

CORES=detectCores()

ann450k <- getAnnotation(IlluminaHumanMethylation450kanno.ilmn12.hg19)
myann <- data.frame(ann450k[,c("UCSC_RefGene_Name","Regulatory_Feature_Group")])
promoters <- grep("Prom",myann$Regulatory_Feature_Group)

# make a list of probes that belong to a gene
gn <- unique(unlist(strsplit( myann$UCSC_RefGene_Name ,";")))
gnl <- strsplit( myann$UCSC_RefGene_Name ,";")
gnl <- mclapply(gnl,unique,mc.cores=CORES)
myann$UCSC_RefGene_Name <- gnl

gnl <- gnl[which(lapply(gnl,length)>0)]

if ( ! file.exists("sets.Rds") ) {

  l <- mclapply(1:nrow(myann), function(i) {
    a <- myann[i,]
    len <- length(a[[1]][[1]])
    probe <- rep(rownames(a),len)
    genes <- a[[1]][[1]]
    data.frame(genes,probe)
  },mc.cores=CORES)

  df <- do.call(rbind,l)

  sets <- mclapply(X=gn, FUN=function(g) {
    df[which(df$genes == g),2]
  } , mc.cores=CORES/2)

  names(sets) <- gn

  saveRDS(object=sets,file="sets.Rds")

} else {

  sets <- readRDS("sets.Rds")

}

```

# DM3 - association of Hcy with gene methylation

It is thought that high plasma homocysteine can inhibit DNA methylation.
Let's see whether that is the case, and which genes are most affected.
This analysis is conducted at the whole gene level as well as on the level of promoters.

## Read in and inspect the data

```{r,dma3a}

dm <- read.table("dma3a.tsv.gz")
dm <- dm[,4:9]
dm <- merge(myann,dm,by=0)
rownames(dm) <- dm[,1]
dm[,1] = NULL

# testing
#dm <- head(dm,40000)

head(dm,50) %>%
  kbl(caption = "Top significant genes with limma") %>%
  kable_paper("hover", full_width = F)

# trim down the input dataset
dm <- dm[,c("UCSC_RefGene_Name","t")]

# histogram of t values
hist(dm$t,breaks=seq(from=-6,to=6,by=1))

```

## Whole gene analysis

```{r,wg}
# set cores to used for parallel execution
CORES= detectCores()

calc_sc <- function(dm) {
  gn <- unique(unlist(strsplit( dm$UCSC_RefGene_Name ,";")))
  gnl <- strsplit( dm$UCSC_RefGene_Name ,";")
  gnl <- mclapply(gnl,unique,mc.cores=CORES)
  dm$UCSC_RefGene_Name <- gnl
  l <- mclapply(1:nrow(dm), function(i) {
    a <- dm[i,]
    len <- length(a[[1]][[1]])
    tvals <- as.numeric(rep(a[2],len))
    genes <- a[[1]][[1]]
    data.frame(genes,tvals)
  },mc.cores=CORES)
  df <- do.call(rbind,l)
  gme_res <- mclapply( 1:length(gn), function(i) {
    g <- gn[i]
    tstats <- df[which(df$genes==g),"tvals"] 
    myn <- length(tstats)
    mymean <- mean(tstats)
    mymedian <- median(tstats)
    wtselfcont <- wilcox.test(tstats)
    res <- c("gene"=g,"nprobes"=myn,"mean"=mymean,"median"=mymedian,
      "p-value(sc)"=wtselfcont$p.value)
  } , mc.cores=CORES )
  gme_res_df <- do.call(rbind, gme_res)
  rownames(gme_res_df) <- gme_res_df[,1]
  gme_res_df <- gme_res_df[,-1]
  tmp <- apply(gme_res_df,2,as.numeric)
  rownames(tmp) <- rownames(gme_res_df)
  gme_res_df <- as.data.frame(tmp)
  gme_res_df$sig <- -log10(gme_res_df[,4])
  gme_res_df <- gme_res_df[order(-gme_res_df$sig),]
  gme_res_df$`fdr(sc)` <- p.adjust(gme_res_df$`p-value(sc)`)
  out <- list("df"=df,"gme_res_df"=gme_res_df)
  return(out)
}

tic()
gme_res_wholegene <- calc_sc(dm)
time2 <- toc() #38 44 41 40 44 41 40
time2
df <- gme_res_wholegene[[1]]
res <- gme_res_wholegene[[2]]
write.table(res,file="gmea_wholegene.tsv")

```

## Table of top results.

```{r, toptable}

head(res,50) %>%
  kbl(caption = "Top significant genes with GMEA") %>%
  kable_paper("hover", full_width = F)

```

## Volcano plots

```{r,volcano1}

# volcano selfcont
sig <- subset(res,`fdr(sc)` < 0.05)
plot(res$median , -log10(res$`p-value(sc)`) ,
  xlab="effect size (mean t-stat)", ylab="-log10(p-value)",
  pch=19, cex=0.5, col="gray",main="self contained test")
grid()
points(sig$median , -log10(sig$`p-value(sc)`) ,
  pch=19, cex=0.5, col="red")

```

## Boxplots

Boxplots smallest pvalue.

```{r,box1}

par(mfrow=c(1,2))
n=50
# self contained

gs <- head(rownames(res),50)
tstats <- lapply(gs, function(g) {
  df[which(df$genes==g),"tvals"]
})
names(tstats) <- gs

tstats <- tstats[order(unlist(lapply(tstats,median)))]
boxplot(tstats,horizontal=TRUE,las=1,
  main="smallest p-val(selfcont)",cex.axis=0.6,
  xlab="t-statistic")
grid()

n=50
# effect size median
sig <- subset(res,`fdr(sc)` < 0.05)
gs <- head(rownames(sig[order(-abs(sig$median)),]),n)

if ( length(gs) >2 ) {
  tstats <- lapply(gs, function(g) {
    df[which(df$genes==g),"tvals"]
  })
  names(tstats) <- gs
  tstats <- tstats[order(unlist(lapply(tstats,median)))]
  boxplot(tstats,horizontal=TRUE,las=1,
    main="biggest effect size(median)",cex.axis=0.6,
    xlab="t-statistic")
  grid()
} else {
  plot(1)
  mtext("too few significant genes found")
}

```

## Mitch

```{r,dm1_mitch}

dmscore <- data.frame( res$median * res$sig)

rownames(dmscore) <- rownames(res)

colnames(dmscore) <- "metric"

if ( ! file.exists("ReactomePathways.gmt") ) {
  download.file("https://reactome.org/download/current/ReactomePathways.gmt.zip",
    destfile="ReactomePathways.gmt.zip")
  unzip("ReactomePathways.gmt.zip")
}
file.info("ReactomePathways.gmt")
genesets <- gmt_import("ReactomePathways.gmt")
length(genesets)

mres <- mitch_calc(x=dmscore, genesets=genesets, priority="effect")

head(mres$enrichment_result,20) %>%
  kbl(caption = "Top enriched gene sets with GMEA-Mitch") %>%
  kable_paper("hover", full_width = F)

mitch_report(mres,outfile="dma3a_mitch.html",overwrite=TRUE)

```

## Using CAMERA

```{r,cam_func}

stat <- dm$t
names(stat) <- rownames(dm)

# slow
sets2 <- head(sets,1000)
tic()
cres <- cameraPR(statistic=stat, index=sets2, use.ranks = FALSE, inter.gene.cor=0.01, sort = TRUE)
toc()

# fast
CORES=detectCores()
#sets <- head(sets,100000)
sets_split <- split(sets,1:CORES)

tic()
cres <- mclapply(X=sets_split, function(mysets) {
  cameraPR(statistic=stat, index=mysets, use.ranks = FALSE, inter.gene.cor=0.01, sort = TRUE)
}, mc.cores=CORES)
cres <- do.call(rbind,cres)
rownames(cres) <- sapply(strsplit(rownames(cres),"\\."),"[[",2)
cres <- cres[order(cres$PValue),]
g <- rownames(cres)[1]
m <- mclapply(rownames(cres) , function(g) {
  probes <- sets[[g]]
  scores <- stat[which(names(stat) %in% probes)]
  mymedian <- median(scores)
  mymean <- mean(scores)
  c(mymean,mymedian)
},mc.cores=CORES)
mdf <- do.call(rbind,m)
colnames(mdf) <- c("mean","median")
cres <- cbind(mdf,cres)
toc()

head(subset(cres,Direction=="Up"),25)
head(subset(cres,Direction=="Down"),25)

head(cres,50) %>%
  kbl(caption = "Top significant genes with CAMERA") %>%
  kable_paper("hover", full_width = F)

```

## CAMERA + pathway analysis

There's no ES and the p-value is biased, so not sure about this

## Promoter analysis only

```{r,promo}

dm <- read.table("dma3a.tsv.gz")
dm <- dm[,4:9]
dm <- merge(myann,dm,by=0)
rownames(dm) <- dm[,1]
dm[,1] = NULL
dm <- dm[grep("Promoter_Associated",dm$Regulatory_Feature_Group),]

head(dm,50) %>%
  kbl(caption = "Top significant genes with limma") %>%
  kable_paper("hover", full_width = F)

# trim down the input dataset
dm <- dm[,c("UCSC_RefGene_Name","t")]

# histogram of t values
hist(dm$t,breaks=seq(from=-6,to=6,by=1))

```

```{r,promo2}

# set cores to used for parallel execution

tic()
gme_res_promoter <- calc_sc(dm)
time2 <- toc() #15.7 20.2 20.3 17.5 17.9 17.7 20.6 
time2

df <- gme_res_promoter[[1]]
res <- gme_res_promoter[[2]] 
write.table(res ,file="gmea_promo.tsv")

```

## Table of top results.

```{r, toptable_promoter}

head(res,50) %>%
  kbl(caption = "Top significant genes with GMEA") %>%
  kable_paper("hover", full_width = F)

```

## Volcano plots

```{r,volcano1_promoter}

# volcano selfcont
sig <- subset(res,`fdr(sc)` < 0.05)
plot(res$median , -log10(res$`p-value(sc)`) ,
  xlab="effect size (mean t-stat)", ylab="-log10(p-value)",
  pch=19, cex=0.5, col="gray",main="self contained test")
grid()
points(sig$median , -log10(sig$`p-value(sc)`) ,
  pch=19, cex=0.5, col="red")

```

## Boxplots

Boxplots smallest pvalue.

```{r,box1_promoter}

par(mfrow=c(1,2))
n=50
# self contained

gs <- head(rownames(res),50)
tstats <- lapply(gs, function(g) {
  df[which(df$genes==g),"tvals"]
})
names(tstats) <- gs

tstats <- tstats[order(unlist(lapply(tstats,median)))]
boxplot(tstats,horizontal=TRUE,las=1,
  main="smallest p-val(selfcont)",cex.axis=0.6,
  xlab="t-statistic")
grid()

n=50
# effect size median
sig <- subset(res,`fdr(sc)` < 0.05)
gs <- head(rownames(sig[order(-abs(sig$median)),]),n)

if ( length(gs) >2 ) {
  tstats <- lapply(gs, function(g) {
    df[which(df$genes==g),"tvals"]
  })
  names(tstats) <- gs
  tstats <- tstats[order(unlist(lapply(tstats,median)))]
  boxplot(tstats,horizontal=TRUE,las=1,
    main="biggest effect size(median)",cex.axis=0.6,
    xlab="t-statistic")
  grid()
} else {
  plot(1)
  mtext("too few significant genes found")
}


```

## Mitch

```{r,dm1_mitch_promo}

dmscore <- data.frame( res$median * res$sig)

rownames(dmscore) <- rownames(res)

colnames(dmscore) <- "metric"

genesets <- gmt_import("ReactomePathways.gmt")

mres <- mitch_calc(x=dmscore, genesets=genesets,priority="effect")

head(mres$enrichment_result,20) %>%
  kbl(caption = "Top enriched gene sets with GMEA-Mitch (promoter only)") %>%
  kable_paper("hover", full_width = F)

mitch_report(mres,outfile="dma3a_mitch_promo.html",overwrite=TRUE)

```

## Session Information

For reproducibility.

```{r,sessioninfo}

sessionInfo()

```
