---
title: "Novakovic et al 2019 reanalysis"
author: "The methylation in ART meta-analysis group"
date: "`r Sys.Date()`"
output:
  html_document:
    fig_width: 7
    fig_height: 7
  toc: true
theme: cosmo
---

```{r,packages}

source("meth_functions.R")

```

# Obtaining array annotations

```{r,annotation}

anno <- getAnnotation(IlluminaHumanMethylationEPICanno.ilm10b4.hg19)
myann <- data.frame(anno[,c("UCSC_RefGene_Name","Regulatory_Feature_Group","Islands_Name","Relation_to_Island")])
promoters <- grep("Prom",myann$Regulatory_Feature_Group)

```

## Introduction

This report is the reanalysis of first analysed DNA methylation data by
[Novakovic et al (2019)] (https://doi.org/10.1038/s41467-019-11929-9)

In this study,  DNA methylation status was generated for 149 neonatal (84♀ 65♂) and 158 adult (87♀ 71♂) 
ART-conceived individuals and for 58 neonatal (37♀, 21♂) and 75 adult (51♀, 24♂) non-ART conceived individuals.


```{r,download}

WORKING_DIR="GSE131433"
ARRAY_DATA="GSE131433_RAW.tar"
DEST=paste(WORKING_DIR,"/",ARRAY_DATA,sep="")
IDAT="GSE131433/IDAT"
options(timeout=1000000)

if(!dir.exists(WORKING_DIR)){
  dir.create(WORKING_DIR)
}

if(!file.exists(DEST)){
  download.file("https://www.ncbi.nlm.nih.gov/geo/download/?acc=GSE131433&format=file",
  #  download.file("ftp://ftp.ncbi.nlm.nih.gov/geo/series/GSE131nnn/GSE131433/suppl/GSE131433_RAW.tar",
    destfile=DEST)
  untar(exdir = "IDAT", tarfile = DEST)
}

SERIES_MATRIX=paste(WORKING_DIR,"/","GSE131433_series_matrix.txt.gz",sep="")
if(!file.exists(SERIES_MATRIX)){
  download.file("ftp://ftp.ncbi.nlm.nih.gov/geo/series/GSE131nnn/GSE131433/matrix/GSE131433_series_matrix.txt.gz", 
    destfile=SERIES_MATRIX)
}
gse <- getGEO(filename=SERIES_MATRIX)

baseDir <- "."
sample_metadata <- pData(phenoData(gse))
targets <- sample_metadata
targets <- subset(targets,`time of collection:ch1`=="birth")
targets$Basename <- head( gsub("_Grn.idat.gz","",list.files("IDAT",full.names=TRUE,pattern="Grn")) , 207 )
head(targets)

rgSet <- read.metharray.exp(targets = targets)
rgSet

```

```{r,norm,fig.cap="Figure 1. Normalisation of bead-array data with SWAN."}

mSet <- preprocessRaw(rgSet)
mSetSw <- SWAN(mSet,verbose=FALSE)
par(mfrow=c(1,2), cex=0.8)
densityByProbeType(mSet[,1], main = "Raw")
densityByProbeType(mSetSw[,1], main = "SWAN")

```


```{r,filterprobes}

# include sex chromosomes
detP <- detectionP(rgSet)
keep <- rowSums(detP < 0.01) == ncol(rgSet)
mSetSw <- mSetSw[keep,]

# exclude SNP probes
mSetSw <- mapToGenome(mSetSw)
mSetSw_nosnp <- dropLociWithSnps(mSetSw)

# exclude sex chromosomes
xyprobes <- anno$Name[anno$chr %in% c("chrX","chrY")]
mSetFlt <- mSetSw[which(!rownames(mSetSw) %in% xyprobes),]

```

```{r,beta_m_vals}

# include sex chromosomes
meth <- getMeth(mSetSw)
unmeth <- getUnmeth(mSetSw)
Mval <- log2((meth + 100)/(unmeth + 100))
beta <- getBeta(mSetSw)
dim(Mval)

# exclude sex chromosomes
meth <- getMeth(mSetFlt)
unmeth <- getUnmeth(mSetFlt)
Mval_flt <- log2((meth + 100)/(unmeth + 100))
beta_flt <- getBeta(mSetFlt)
dim(Mval_flt)

```

# MDS analysis

```{r,scree,fig.cap="Figure 2. Screeplot shows contribution of top principal components to the overall variation in the experiment."}

par(mfrow=c(2,1))
myscree(Mval,main="incl sex chr")
myscree(Mval_flt,main="excl sex chr")

```

#MDS plot 1

```{r,mds1,fig.width = 8 ,fig.height = 8,fig.cap="Figure 3. MDS plot coloured by ART classification."}

colnames(targets) <- gsub(" ","_",colnames(targets))
colnames(targets) <- gsub(":","_",colnames(targets))
targets$sex<-factor(targets$`gender_ch1`)
targets$art<-factor(targets$`art_subtype`)
sample_groups <- factor(targets$art)
colour_palette=brewer.pal(n = length(levels(sample_groups)), name = "Paired")
colours <- colour_palette[as.integer(factor(targets$art))]
plot(1,axes = FALSE,xlab="",ylab="",main="MDS by ART type")
legend("center",legend=levels(sample_groups),pch=16,cex=1.2,col=colour_palette)
mydist <- plotMDS(Mval, labels=targets$Sample_Name,col=colours,main="sex chromosomes included")
mydist_flt <- plotMDS(Mval_flt, labels=targets$Sample_Name,col=colours,main="sex chromosomes excluded")
colnames(mydist_flt@.Data[[5]])  <- sapply(strsplit(colnames(mydist_flt@.Data[[5]]),"_"),"[[",1)
plotMDS(mydist_flt, labels=targets$Sample_Name,col=colours,main="sex chromosomes excluded")

```

#save data object

```{r,save}

save.image("Novakovic.Rdata")

```


#Differential analysis

1. non-ART Vs Fresh embryo

2. non-ART Vs Frozen embryo

3. non-ART Vs GIFT

4. non-ART Vs FX

5. GIFT Vs FX

6. non-ART Vs ART

7. Fresh Vs Frozen

### non-ART Vs Fresh embryo

```{r,dm1}

# sex chromosomes excluded
beta <- beta_flt
Mval <- Mval_flt
colnames(beta) <- sapply(strsplit(colnames(beta),"_"),"[[",1)
colnames(Mval) <- sapply(strsplit(colnames(Mval),"_"),"[[",1)

#targets$Basename <- gsub("IDAT/","",targets$Basename)  
birth<-targets[which(targets$`time_of_collection`=="birth"),]
birth$Basename <- sapply(strsplit(birth$Basename,"/"),"[[",3)
birth$Basename <- sapply(strsplit(birth$Basename,"_"),"[[",1)

samplesheet <- subset(birth,art_subtype_ch1=="non-ART"|art_subtype_ch1=="Fresh embryo")
groups <- factor(samplesheet$art_subtype,levels=c("non-ART","Fresh embryo"))
sex <- factor(samplesheet$sex,levels=c("M","F"))
Mvals<-Mval[,colnames(Mval)%in% samplesheet$Basename]
betas<-beta[,colnames(beta)%in% samplesheet$Basename]
dim(Mvals)
dim(betas)

top_nat_vs_fh <- dm_analysis(samplesheet=samplesheet,
  sex=sex,groups=groups,mx=Mvals,name="top_nat_vs_fh",
  myann=myann ,beta= betas)
head(top_nat_vs_fh$dma,10)
head(top_nat_vs_fh$dmr,10)
top_nat_vs_fh$comp
top_nat_vs_fh$cgi
names(top_nat_vs_fh)
head(top_nat_vs_fh$confects$table$name)

saveRDS(top_nat_vs_fh,file="novakovic_nat_vs_fh.rds")

```


### non-ART Vs Frozen embryo

```{r,dm2}

samplesheet<-subset(birth,art_subtype_ch1=="non-ART"|art_subtype_ch1=="Frozen embryo")
groups <- factor(samplesheet$art_subtype,levels=c("non-ART","Frozen embryo"))
sex <- factor(samplesheet$sex,levels=c("M","F"))
Mvals<-Mval[,colnames(Mval)%in% samplesheet$Basename]
betas<-beta[,colnames(beta)%in% samplesheet$Basename]

top_nat_vs_fz <- dm_analysis(samplesheet=samplesheet,
  sex=sex,groups=groups,mx=Mvals,name="top_nat_vs_fz",
  myann=myann ,beta= betas)

head(top_nat_vs_fz$dma,10)
head(top_nat_vs_fz$dmr,10)
top_nat_vs_fz$comp
top_nat_vs_fz$cgi
names(top_nat_vs_fz)
head(top_nat_vs_fz$confects$table$name)

saveRDS(top_nat_vs_fz,file="novakovic_nat_vs_fz.rds")

```

### non-ART Vs GIFT

```{r,dm3}

samplesheet<-subset(birth,art_subtype_ch1=="non-ART"|art_subtype_ch1=="GIFT")
groups <- factor(samplesheet$art_subtype,levels=c("non-ART","GIFT"))
sex <- factor(samplesheet$sex,levels=c("M","F"))
Mvals<-Mval[,colnames(Mval)%in% samplesheet$Basename]
betas<-beta[,colnames(beta)%in% samplesheet$Basename]

top_nat_vs_GIFT <- dm_analysis(samplesheet=samplesheet,
sex=sex,groups=groups,mx=Mvals,name="top_nat_vs_GIFT",
myann=myann ,beta= betas)

head(top_nat_vs_GIFT$dma,10)
head(top_nat_vs_GIFT$dmr,10)

top_nat_vs_GIFT$comp
top_nat_vs_GIFT$cgi
names(top_nat_vs_GIFT)
head(top_nat_vs_GIFT$confects$table$name)

saveRDS(top_nat_vs_GIFT,file="novakovic_nat_vs_GIFT.rds")

```

### non-ART Vs FX

```{r,dm4}

samplesheet <- subset(birth,art_subtype_ch1=="non-ART"|art_subtype_ch1=="Frozen embryo" | art_subtype_ch1=="Fresh embryo")
samplesheet$art_subtype <- gsub("Fresh embryo","FX",samplesheet$art_subtype)
samplesheet$art_subtype <- gsub("Frozen embryo","FX",samplesheet$art_subtype)
groups <- factor(samplesheet$art_subtype,levels=c("non-ART","FX"))

sex <- factor(samplesheet$sex,levels=c("M","F"))
Mvals<-Mval[,colnames(Mval)%in% samplesheet$Basename]
betas<-beta[,colnames(beta)%in% samplesheet$Basename]

top_nat_vs_FX <- dm_analysis(samplesheet=samplesheet,
  sex=sex,groups=groups,mx=Mvals,name="top_nat_vs_FX",
  myann=myann ,beta= betas)

head(top_nat_vs_FX$dma,10)
head(top_nat_vs_FX$dmr,10)

top_nat_vs_FX$comp
top_nat_vs_FX$cgi
names(top_nat_vs_FX)
head(top_nat_vs_FX$confects$table$name)

saveRDS(top_nat_vs_FX,file="novakovic_nat_vs_FX.rds")

```

### GIFT Vs FX

```{r,dm5}

samplesheet <- subset(birth,art_subtype_ch1=="GIFT"|art_subtype_ch1=="Frozen embryo" | art_subtype_ch1=="Fresh embryo")
samplesheet$art_subtype <- gsub("Fresh embryo","FX",samplesheet$art_subtype)
samplesheet$art_subtype <- gsub("Frozen embryo","FX",samplesheet$art_subtype)
groups <- factor(samplesheet$art_subtype,levels=c("GIFT","FX"))
sex <- factor(samplesheet$sex,levels=c("M","F"))
Mvals<-Mval[,colnames(Mval)%in% samplesheet$Basename]
betas<-beta[,colnames(beta)%in% samplesheet$Basename]

top_GIFT_vs_FX <- dm_analysis(samplesheet=samplesheet,
sex=sex,groups=groups,mx=Mvals,name="top_GIFT_vs_FX",
myann=myann ,beta= betas)
head(top_GIFT_vs_FX$dma,10)
head(top_GIFT_vs_FX$dmr,10)
top_GIFT_vs_FX$comp
top_GIFT_vs_FX$cgi
names(top_GIFT_vs_FX)
head(top_GIFT_vs_FX$confects$table$name)

saveRDS(top_GIFT_vs_FX,file="novakovic_GIFT_vs_FX.rds")

```

### non-ART Vs ART

```{r,dm6}

samplesheet<-birth
groups <- factor(samplesheet$art_status,levels=c("non-ART","ART"))
sex <- factor(samplesheet$sex,levels=c("M","F"))
Mvals<-Mval[,colnames(Mval)%in% samplesheet$Basename]
betas<-beta[,colnames(beta)%in% samplesheet$Basename]

top_nat_vs_ART <- dm_analysis(samplesheet=samplesheet,
sex=sex,groups=groups,mx=Mvals,name="top_nat_vs_ART",
myann=myann ,beta= betas)
head(top_nat_vs_ART$dma,10)
head(top_nat_vs_ART$dmr,10)
top_nat_vs_ART$comp
top_nat_vs_ART$cgi
names(top_nat_vs_ART)
head(top_nat_vs_ART$confects$table$name)

saveRDS(top_nat_vs_ART,file="novakovic_nat_vs_ART.rds")

```


### Fresh embryo Vs Frozen embryo

```{r, dm7}

samplesheet <- subset(birth,art_subtype_ch1=="Fresh embryo"|art_subtype_ch1=="Frozen embryo")
groups <- factor(samplesheet$art_subtype,levels=c("Fresh embryo","Frozen embryo"))
sex <- factor(samplesheet$sex,levels=c("M","F"))
Mvals<-Mval[,colnames(Mval)%in% samplesheet$Basename]
betas<-beta[,colnames(beta)%in% samplesheet$Basename]

top_fh_vs_fz <- dm_analysis(samplesheet=samplesheet,
sex=sex,groups=groups,mx=Mvals,name="top_fh_vs_fz",
myann=myann ,beta= betas)
head(top_fh_vs_fz$dma,10)
head(top_fh_vs_fz$dmr,10)
top_fh_vs_fz$comp
top_fh_vs_fz$cgi
names(top_fh_vs_fz)
head(top_fh_vs_fz$confects$table$name)

saveRDS(top_fh_vs_fz,file="novakovic_fh_vs_fz.rds")

```

## SessionInfo

```{r,session}

sessionInfo()

```
