---
title: "Testing gene methylation enrichment analysis approaches using simulated data"
author: "The GMEA team"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    fig_width: 7
    fig_height: 7
theme: cosmo
---

## Introduction

In this report we establish a new method for generating simulated data with known ground truth.
This will be used to test different gene methylation enrichment approaches systematically.

The general steps are:

1. Import GSE158422 data corresponding to control (non-tumour tissue).

2. From the 37 samples, create two groups of 18 samples.
One of these will be considered "control" and the other "case".

3. Create random gene sets that have similar sized to Reactome pathways.

3. Some gene sets will be selected to be differentially methylated.
Half of these will be hypermethylated and the others will be hypomethylated.

4. The changes will be incorporated into the "case" samples.

5. The enrichment analysis will be conducted.

6. The accuracy will be calculated.

```{r,packages}

suppressPackageStartupMessages({
  library("stringi")
  library("limma")
  library("missMethyl")
  library("IlluminaHumanMethylation450kmanifest")
  library("IlluminaHumanMethylationEPICanno.ilm10b4.hg19")
  library("HGNChelper")
  library('org.Hs.eg.db')
  library("psych")
  library("mitch")
  library("kableExtra")
  library("rslurm")
})

source("GSE158422_simulate_functions.R")

# optimised for 128 GB sever with 32 threads
CORES=6

```

## Load data

* annotations

* probe sets

* gene sets

* design matrix

* mval matrix

```{r,annotation}

# get probe-gene mapping
anno <- getAnnotation(IlluminaHumanMethylationEPICanno.ilm10b4.hg19)
myann <- data.frame(anno[,c("UCSC_RefGene_Name","UCSC_RefGene_Group","Islands_Name","Relation_to_Island")])
gp <- myann[,"UCSC_RefGene_Name",drop=FALSE]
gp2 <- strsplit(gp$UCSC_RefGene_Name,";")
names(gp2) <- rownames(gp)
gp2 <- lapply(gp2,unique)
gt <- stack(gp2)
colnames(gt) <- c("gene","probe")
dim(gt)
str(gt)
head(gt)

#new.hgnc.table <- getCurrentHumanMap()
#new.hgnc.table <- readRDS("new.hgnc.table.rds")
#fix <- checkGeneSymbols(gt$gene,map=new.hgnc.table)
#fix2 <- fix[which(fix$x != fix$Suggested.Symbol),]
#length(unique(fix2$x))
#gt$gene <- fix$Suggested.Symbol

if (!file.exists("GSE158422_design.rds")) {
  download.file("https://ziemann-lab.net/public/gmea_prototype/GSE158422_design.rds", "GSE158422_design.rds")
}
design <- readRDS("GSE158422_design.rds")

if (!file.exists("GSE158422_design.rds")) {
  options(timeout=10000000000000)
  download.file("https://ziemann-lab.net/public/gmea_prototype/GSE158422_mx.rds","GSE158422_mx.rds")
}
mval <- readRDS("GSE158422_mx.rds")

```

# Gene set database

We could use Reactome pathways, however these have a lot of overlapping sets, which could cause inflated
false positives.
After some testing, I found that the gene set size strongly impacts the accuracy.

For reference, Reactome sets have a mean of 48 and median of 15, while MSigDB has a median of 47 and mean of 116.

Therefore, a reasonable analysis would include small (20), medium (50) and large (100) sets.

To reflect coverage of Reactome and other pathway databases, only half the genes will be included in sets.

```{r,gset_prep}

gene_catalog <- unique(gt$gene)

# set gene set size here
lengths <- rep(20,1000)

gsets <- randomGeneSets(gene_catalog,lengths,seed=100)

```

# Incorporate changes

TODO: to incorporate changes to case samples.

Need to figure out what magnitude to change.
Will refer to the cancer/normal comparison.

Select genes and probes to alter.

```{r,incorp}

seed=100
frac_genes=0.5
frac_probes=0.5
delta=1
nsamples=10
normal_mval <- mval[,(1:(ncol(mval)/2)*2)]

```

# Run analyses

Set assumptions.

```{r,set}

num_dm_sets=50
sims=100
groupsizes=c(3,5,8,12)
deltas=c(0.1,0.2,0.3,0.4,0.5)
#groupsizes=12
#deltas=0.5

params <- expand.grid("groupsizes"=groupsizes,"deltas"=deltas,"seed"=seq(100,sims*100,100))

colnames(params) <- c("groupsize","delta","seed")

str(params)

params <- params[order(params$groupsize),]

```

## GSA

Using GSAmeth of significant probes.

```{r,rungsa}

sopt1 <- list(time = '6:00:00', mem="16G")

sjob <- slurm_apply(simgsa, params, jobname = 'gsa',
  nodes = 50, cpus_per_node = 1, submit = TRUE, slurm_options = sopt1,
  genesetdatabase=gsets, myann=myann, mval=normal_mval, frac_genes=0.5,
  frac_probes=0.5, num_dm_sets=num_dm_sets,
  pkgs = rev(.packages()),
  global_objects = c("randomGeneSets", "runlimma","incorp_dm", "F1", "gsets") )

gres <- get_slurm_out(sjob, outtype = 'raw', wait = TRUE)

gres <- do.call(rbind,gres)

gres <- data.frame(params,gres)

str(gres)

```


## LAT

Limma Average t-test.

```{r,runlat}

sopt1 <- list(time = '6:00:00', mem="32G")

sjob <- slurm_apply(simlac, params, jobname = 'lat',
  nodes = 50, cpus_per_node = 2, submit = TRUE, slurm_options = sopt1,
  genesetdatabase=gsets, myann=myann, mval=normal_mval, frac_genes=0.5,
  frac_probes=0.5, num_dm_sets=num_dm_sets,
  pkgs = rev(.packages()),
  global_objects = c("randomGeneSets", "runlimma","incorp_dm", "ttenrich","F1", "gsets","gt") )

latres <- get_slurm_out(sjob, outtype = 'raw', wait = TRUE)

latres <- do.call(rbind,latres)

latres <- data.frame(params,latres)

str(latres)

```

## LTT

Limma top t-test.

```{r,runltt}

sopt1 <- list(time = '6:00:00', mem="32G")

sjob <- slurm_apply(simlactop, params, jobname = 'ltt',
  nodes = 50, cpus_per_node = 2, submit = TRUE, slurm_options = sopt1,
  genesetdatabase=gsets, myann=myann, mval=normal_mval, frac_genes=0.5,
  frac_probes=0.5, num_dm_sets=num_dm_sets,
  pkgs = rev(.packages()),
  global_objects = c("randomGeneSets", "runlimma","incorp_dm", "ttenrich","F1", "gsets","gt") )

lttres <- get_slurm_out(sjob, outtype = 'raw', wait = TRUE)

lttres <- do.call(rbind,lttres)

lttres <- data.frame(params,lttres)

str(lttres)

```

## LAW

Limma average wilcox.

```{r,runlaw}

sopt1 <- list(time = '6:00:00', mem="32G")

sjob <- slurm_apply(simnlac, params, jobname = 'law',
  nodes = 50, cpus_per_node = 2, submit = TRUE, slurm_options = sopt1,
  genesetdatabase=gsets, myann=myann, mval=normal_mval, frac_genes=0.5,
  frac_probes=0.5, num_dm_sets=num_dm_sets,
  pkgs = rev(.packages()),
  global_objects = c("randomGeneSets", "runlimma","incorp_dm", "wtenrich","F1", "gsets","gt") )

lawres <- get_slurm_out(sjob, outtype = 'raw', wait = TRUE)

lawres <- do.call(rbind,lawres)

lawres <- data.frame(params,lawres)

str(lawres)

```

## LAM

Limma average mitch.

```{r,runlam}

sopt1 <- list(time = '6:00:00', mem="32G")

sjob <- slurm_apply(simlacm, params, jobname = 'lam',
  nodes = 50, cpus_per_node = 2, submit = TRUE, slurm_options = sopt1,
  genesetdatabase=gsets, myann=myann, mval=normal_mval, frac_genes=0.5,
  frac_probes=0.5, num_dm_sets=num_dm_sets,
  pkgs = rev(.packages()),
  global_objects = c("randomGeneSets", "runlimma","incorp_dm", "runmitch","F1", "gsets","gt") )

lamres <- get_slurm_out(sjob, outtype = 'raw', wait = TRUE)

lamres <- do.call(rbind,lamres)

lamres <- data.frame(params,lamres)

str(lamres)

```

## LRM

Limma rank mitch.

```{r,runlrm}

sopt1 <- list(time = '6:00:00', mem="32G")

sjob <- slurm_apply(simlrm, params, jobname = 'lrm',
  nodes = 50, cpus_per_node = 2, submit = TRUE, slurm_options = sopt1,
  genesetdatabase=gsets, myann=myann, mval=normal_mval, frac_genes=0.5,
  frac_probes=0.5, num_dm_sets=num_dm_sets,
  pkgs = rev(.packages()),
  global_objects = c("randomGeneSets", "runlimma","incorp_dm", "runmitch","F1", "gsets","gt") )

lrmres <- get_slurm_out(sjob, outtype = 'raw', wait = TRUE)

lrmres <- do.call(rbind,lrmres)

lrmres <- data.frame(params,lrmres)

str(lrmres)

```

## ALT

Aggregate limma t-test

```{r,runalt}

sopt1 <- list(time = '6:00:00', mem="32G")

sjob <- slurm_apply(simalc, params, jobname = 'alt',
  nodes = 50, cpus_per_node = 2, submit = TRUE, slurm_options = sopt1,
  genesetdatabase=gsets, myann=myann, mval=normal_mval, frac_genes=0.5,
  frac_probes=0.5, num_dm_sets=num_dm_sets,
  pkgs = rev(.packages()),
  global_objects = c("randomGeneSets", "runlimma","incorp_dm", "ttenrich","F1", "gsets","gt") )

altres <- get_slurm_out(sjob, outtype = 'raw', wait = TRUE)

altres <- do.call(rbind,altres)

altres <- data.frame(params,altres)

str(altres)

```

## ALW

Aggregate limma wilcox test

```{r,runalw}

sopt1 <- list(time = '6:00:00', mem="32G")

sjob <- slurm_apply(simnalc, params, jobname = 'alw',
  nodes = 50, cpus_per_node = 2, submit = TRUE, slurm_options = sopt1,
  genesetdatabase=gsets, myann=myann, mval=normal_mval, frac_genes=0.5,
  frac_probes=0.5, num_dm_sets=num_dm_sets,
  pkgs = rev(.packages()),
  global_objects = c("randomGeneSets", "runlimma","incorp_dm", "wtenrich","F1", "gsets","gt") )

alwres <- get_slurm_out(sjob, outtype = 'raw', wait = TRUE)

alwres <- do.call(rbind,alwres)

alwres <- data.frame(params,alwres)

str(alwres)

```

## ALM

Aggregate limma mitch

```{r,runalm}

sopt1 <- list(time = '6:00:00', mem="32G")

sjob <- slurm_apply(simalm, params, jobname = 'alm',
  nodes = 50, cpus_per_node = 2, submit = TRUE, slurm_options = sopt1,
  genesetdatabase=gsets, myann=myann, mval=normal_mval, frac_genes=0.5,
  frac_probes=0.5, num_dm_sets=num_dm_sets,
  pkgs = rev(.packages()),
  global_objects = c("randomGeneSets", "runlimma","incorp_dm", "runmitch","F1", "gsets","gt") )

almres <- get_slurm_out(sjob, outtype = 'raw', wait = TRUE)

almres <- do.call(rbind,almres)

almres <- data.frame(params,almres)

str(almres)

```

## Session information

```{r,save}

save.image("GSE158422_simulate020_rslurm.Rdata")

sessionInfo()

```
