---
title: "Testing gene methylation enrichment analysis approaches using simulated data - summarise data"
author: "The GMEA team"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    fig_width: 7
    fig_height: 7
theme: cosmo
---

## Introduction

Here is an explanation of the different objects:

0. GSA: gres

1. LAT: lacres

2. LTT: lactopres

3. LAW: nlacres

4. LAM: lacmres

5. LRM: lrmres

6. ALT: alcres

7. ALW: nalcres

8. ALM: almres

```{r,packages}

suppressPackageStartupMessages({
  library("stringi")
  library("limma")
  library("missMethyl")
  library("IlluminaHumanMethylation450kmanifest")
  library("IlluminaHumanMethylationEPICanno.ilm10b4.hg19")
  library('org.Hs.eg.db')
  library("psych")
  library("mitch")
  library("kableExtra")
})

```

## Functions

```{r,functions}

F1 <- function(x,y) {
  ( 2 * x * y ) / ( x + y )
}

summarise <- function(x) {
  PREC <- sum(x[,"TP"]) / ( sum(x[,"TP"]) + sum(x[,"FP"]) )
  REC <- sum(x[,"TP"]) / ( sum(x[,"TP"]) + sum(x[,"FN"]) )
  F <- F1(PREC,REC)
  return(c("PRECISION"=PREC,"RECALL"=REC,"F1"=F))
}

```

## Overall summary

Make a set of three barplots which summarise method performance over the range of group size and
delta values, for gene set sizes of 20, 50 and 200.

```{r,summary,fig.width=5,fig.height=4}

mar = c(5.1, 4.1, 5.1, 2.1)

load("GSE158422_simulate020.Rdata")

l1 <- list("GSA"=gres2,"LAT"=lacres2,"LTT"=lactopres2,"LAW"=nlacres2,
  "LAM"=lacmres2, "LRM"=lrmres2, "ALT"=alcres2,"ALW"=nalcres2,"ALM"=almres2)

x1 <- do.call(rbind,lapply(l1,summarise))

x1 %>% kbl(caption="sim 20 res") %>% kable_paper("hover", full_width = F)

cols <- c("white","gray","black")

barplot(t(x1),beside=TRUE,ylim=c(0,1),ylab="index", col=cols, cex.names=0.7, cex.axis=0.7, main="20 genes per set")
grid()

legend("topleft", inset=.1, c("Precision","Recall","F1"), fill=cols, cex=0.7)

pdf("fig1a.pdf",width=5,height=4)
barplot(t(x1),beside=TRUE,ylim=c(0,1),ylab="index", col=cols, cex.names=0.7, cex.axis=0.7, main="20 genes per set")
grid()
legend("topleft", inset=.1, c("Precision","Recall","F1"), fill=cols, cex=0.7)
dev.off()

load("GSE158422_simulate050.Rdata")

l2 <- list("GSA"=gres2,"LAT"=lacres2,"LTT"=lactopres2,"LAW"=nlacres2,
  "LAM"=lacmres2, "LRM"=lrmres2, "ALT"=alcres2,"ALW"=nalcres2,"ALM"=almres2)

x2 <- do.call(rbind,lapply(l2,summarise))

x2 %>% kbl(caption="sim 50 res") %>% kable_paper("hover", full_width = F)

barplot(t(x2),beside=TRUE,ylim=c(0,1),ylab="index", col=cols, cex.names=0.7, cex.axis=0.7, main="50 genes per set")
grid()

pdf("fig1b.pdf",width=5,height=4)
barplot(t(x2),beside=TRUE,ylim=c(0,1),ylab="index", col=cols, cex.names=0.7, cex.axis=0.7, main="50 genes per set")
grid()
dev.off()

load("GSE158422_simulate100.Rdata")

l3 <- list("GSA"=gres2,"LAT"=lacres2,"LTT"=lactopres2,"LAW"=nlacres2,
  "LAM"=lacmres2, "LRM"=lrmres2, "ALT"=alcres2,"ALW"=nalcres2,"ALM"=almres2)

x3 <- do.call(rbind,lapply(l3,summarise))

x3 %>% kbl(caption="sim 100 res") %>% kable_paper("hover", full_width = F)

barplot(t(x3),beside=TRUE,ylim=c(0,1),ylab="index", col=cols, cex.names=0.7, cex.axis=0.7, main="100 genes per set")
grid()

pdf("fig1c.pdf",width=5,height=4)
barplot(t(x3),beside=TRUE,ylim=c(0,1),ylab="index", col=cols, cex.names=0.7, cex.axis=0.7, main="100 genes per set")
grid()
legend("topleft", inset=.1, c("Precision","Recall","F1"), fill=cols, cex=0.7)
dev.off()

```

Make a PNG copy of the data.

```{r,summary_png,fig.width=5,fig.height=4}

png(filename="fig1.png",height=70,width=210,units="mm",res=200)

par(mar = c(2.5, 4.1, 2.1, 0.5))

par(mfrow=c(1,3))

cols <- c("white","gray","black")

barplot(t(x1),beside=TRUE,ylim=c(0,1),ylab="index", col=cols, cex.names=0.75, cex.axis=0.75, main="20 genes per set")
grid()

legend("topleft", inset=.1, c("Precision","Recall","F1"), fill=cols, cex=0.75)

barplot(t(x2),beside=TRUE,ylim=c(0,1),ylab="index", col=cols, cex.names=0.75, cex.axis=0.75, main="50 genes per set")
grid()

barplot(t(x3),beside=TRUE,ylim=c(0,1),ylab="index", col=cols, cex.names=0.75, cex.axis=0.75, main="100 genes per set")
grid()

dev.off()

par(mfrow=c(1,1))

```

## In depth analysis of GSA vs LAM

As GSA appears to be the best-performing existing method for enrichment analysis, it is a good idea
to compare it to LAM which is the best one out of the new methods evaluated here.
The following charts show the performance of GSA and LAM including precision and recall over the
range of group sizes and delta values when the sets contain 20, 50 and 100 genes each.

```{r,020_gsa_vs_lam}

# 20
load("GSE158422_simulate020.Rdata")

par(mfrow=c(2,3))

gres2[,"PREC"] <- gres2[,"TP"] / ( gres2[,"TP"] + gres2[,"FP"] )
gres3p <- do.call(rbind,lapply(groupsizes, function (g) { gres2[params$groupsizes==g,"PREC"] }))
colnames(gres3p) <- deltas
rownames(gres3p) <- groupsizes
gres3p %>% kbl(caption="GSAmeth precision") %>% kable_paper("hover", full_width = F)

gres3r <- do.call(rbind,lapply(groupsizes, function (g) { gres2[params$groupsizes==g,"REC"] }))
colnames(gres3r) <- deltas
rownames(gres3r) <- groupsizes
gres3r %>% kbl(caption="GSAmeth recall") %>% kable_paper("hover", full_width = F)

gres3f <- F1(gres3p,gres3r)
gres3f %>% kbl(caption="GSAmeth F1") %>% kable_paper("hover", full_width = F)

plot(rownames(gres3p),gres3p[,"0.5"],type="b",las=1,ylab="precision",xlab="",
  main="",ylim=c(0,1))
lines(rownames(gres3p),gres3p[,"0.4"],type="b",col="red")
lines(rownames(gres3p),gres3p[,"0.3"],type="b",col="blue")
lines(rownames(gres3p),gres3p[,"0.2"],type="b",col="darkgreen")
grid()

legend("bottomright", legend=c("0.5", "0.4", "0.3", "0.2"),title="delta",
  lty=1, lwd=2, col=c("black", "red", "blue", "darkgreen"), cex=1)

plot(rownames(gres3r),gres3r[,"0.5"],type="b",las=1,ylab="recall",xlab="group size",
  main="GSA",ylim=c(0,1))
lines(rownames(gres3r),gres3r[,"0.4"],type="b",col="red")
lines(rownames(gres3r),gres3r[,"0.3"],type="b",col="blue")
lines(rownames(gres3r),gres3r[,"0.2"],type="b",col="darkgreen")
grid()

plot(rownames(gres3r),gres3f[,"0.5"],type="b",las=1,ylab="F1",xlab="",
  main="",ylim=c(0,1))
lines(rownames(gres3r),gres3f[,"0.4"],type="b",col="red")
lines(rownames(gres3r),gres3f[,"0.3"],type="b",col="blue")
lines(rownames(gres3r),gres3f[,"0.2"],type="b",col="darkgreen")
grid()

lacmres2[,"PREC"] <- lacmres2[,"TP"] / ( lacmres2[,"TP"] + lacmres2[,"FP"] )
lacmres3p <- do.call(rbind,lapply(groupsizes, function (g) { lacmres2[params$groupsizes==g,"PREC"] }))
colnames(lacmres3p) <- deltas
rownames(lacmres3p) <- groupsizes
lacmres3p %>% kbl(caption="LAM precision") %>% kable_paper("hover", full_width = F)

lacmres3r <- do.call(rbind,lapply(groupsizes, function (g) { lacmres2[params$groupsizes==g,"REC"] }))
colnames(lacmres3r) <- deltas
rownames(lacmres3r) <- groupsizes
lacmres3r %>% kbl(caption="LAM recall") %>% kable_paper("hover", full_width = F)

lacmres3f <- F1(lacmres3p,lacmres3r)
lacmres3f %>% kbl(caption="LAM F1") %>% kable_paper("hover", full_width = F)

plot(rownames(lacmres3p),lacmres3p[,"0.5"],type="b",las=1,ylab="precision",xlab="",
  main="",ylim=c(0,1))
lines(rownames(lacmres3p),lacmres3p[,"0.4"],type="b",col="red")
lines(rownames(lacmres3p),lacmres3p[,"0.3"],type="b",col="blue")
lines(rownames(lacmres3p),lacmres3p[,"0.2"],type="b",col="darkgreen")
grid()

legend("bottomright", legend=c("0.5", "0.4", "0.3", "0.2"),title="delta",
  lty=1, lwd=2, col=c("black", "red", "blue", "darkgreen"), cex=1)

plot(rownames(lacmres3r),lacmres3r[,"0.5"],type="b",las=1,ylab="recall",xlab="group size",
  main="LAM",ylim=c(0,1))
lines(rownames(lacmres3r),lacmres3r[,"0.4"],type="b",col="red")
lines(rownames(lacmres3r),lacmres3r[,"0.3"],type="b",col="blue")
lines(rownames(lacmres3r),lacmres3r[,"0.2"],type="b",,col="darkgreen")
grid()

plot(rownames(lacmres3r),lacmres3f[,"0.5"],type="b",las=1,ylab="F1",xlab="",
  main="",ylim=c(0,1))
lines(rownames(lacmres3r),lacmres3f[,"0.4"],type="b",col="red")
lines(rownames(lacmres3r),lacmres3f[,"0.3"],type="b",col="blue")
lines(rownames(lacmres3r),lacmres3f[,"0.2"],type="b",col="darkgreen")
grid()

# 50
load("GSE158422_simulate050.Rdata")

gres2[,"PREC"] <- gres2[,"TP"] / ( gres2[,"TP"] + gres2[,"FP"] )
gres3p <- do.call(rbind,lapply(groupsizes, function (g) { gres2[params$groupsizes==g,"PREC"] }))
colnames(gres3p) <- deltas
rownames(gres3p) <- groupsizes
gres3p %>% kbl(caption="GSAmeth precision") %>% kable_paper("hover", full_width = F)

gres3r <- do.call(rbind,lapply(groupsizes, function (g) { gres2[params$groupsizes==g,"REC"] }))
colnames(gres3r) <- deltas
rownames(gres3r) <- groupsizes
gres3r %>% kbl(caption="GSAmeth recall") %>% kable_paper("hover", full_width = F)

gres3f <- F1(gres3p,gres3r)
gres3f %>% kbl(caption="GSAmeth F1") %>% kable_paper("hover", full_width = F)

plot(rownames(gres3p),gres3p[,"0.5"],type="b",las=1,ylab="precision",xlab="",
  main="",ylim=c(0,1))
lines(rownames(gres3p),gres3p[,"0.4"],type="b",col="red")
lines(rownames(gres3p),gres3p[,"0.3"],type="b",col="blue")
lines(rownames(gres3p),gres3p[,"0.2"],type="b",col="darkgreen")
grid()

legend("bottomright", legend=c("0.5", "0.4", "0.3", "0.2"),title="delta",
  lty=1, lwd=2, col=c("black", "red", "blue", "darkgreen"), cex=1)

plot(rownames(gres3r),gres3r[,"0.5"],type="b",las=1,ylab="recall",xlab="group size",
  main="GSA",ylim=c(0,1))
lines(rownames(gres3r),gres3r[,"0.4"],type="b",col="red")
lines(rownames(gres3r),gres3r[,"0.3"],type="b",col="blue")
lines(rownames(gres3r),gres3r[,"0.2"],type="b",col="darkgreen")
grid()

plot(rownames(gres3r),gres3f[,"0.5"],type="b",las=1,ylab="F1",xlab="",
  main="",ylim=c(0,1))
lines(rownames(gres3r),gres3f[,"0.4"],type="b",col="red")
lines(rownames(gres3r),gres3f[,"0.3"],type="b",col="blue")
lines(rownames(gres3r),gres3f[,"0.2"],type="b",col="darkgreen")
grid()

lacmres2[,"PREC"] <- lacmres2[,"TP"] / ( lacmres2[,"TP"] + lacmres2[,"FP"] )
lacmres3p <- do.call(rbind,lapply(groupsizes, function (g) { lacmres2[params$groupsizes==g,"PREC"] }))
colnames(lacmres3p) <- deltas
rownames(lacmres3p) <- groupsizes
lacmres3p %>% kbl(caption="LAM precision") %>% kable_paper("hover", full_width = F)

lacmres3r <- do.call(rbind,lapply(groupsizes, function (g) { lacmres2[params$groupsizes==g,"REC"] }))
colnames(lacmres3r) <- deltas
rownames(lacmres3r) <- groupsizes
lacmres3r %>% kbl(caption="LAM recall") %>% kable_paper("hover", full_width = F)

lacmres3f <- F1(lacmres3p,lacmres3r)
lacmres3f %>% kbl(caption="LAM F1") %>% kable_paper("hover", full_width = F)

plot(rownames(lacmres3p),lacmres3p[,"0.5"],type="b",las=1,ylab="precision",xlab="",
  main="",ylim=c(0,1))
lines(rownames(lacmres3p),lacmres3p[,"0.4"],type="b",col="red")
lines(rownames(lacmres3p),lacmres3p[,"0.3"],type="b",col="blue")
lines(rownames(lacmres3p),lacmres3p[,"0.2"],type="b",col="darkgreen")
grid()

legend("bottomright", legend=c("0.5", "0.4", "0.3", "0.2"),title="delta",
  lty=1, lwd=2, col=c("black", "red", "blue", "darkgreen"), cex=1)

plot(rownames(lacmres3r),lacmres3r[,"0.5"],type="b",las=1,ylab="recall",xlab="group size",
  main="LAM",ylim=c(0,1))
lines(rownames(lacmres3r),lacmres3r[,"0.4"],type="b",col="red")
lines(rownames(lacmres3r),lacmres3r[,"0.3"],type="b",col="blue")
lines(rownames(lacmres3r),lacmres3r[,"0.2"],type="b",,col="darkgreen")
grid()

plot(rownames(lacmres3r),lacmres3f[,"0.5"],type="b",las=1,ylab="F1",xlab="",
  main="",ylim=c(0,1))
lines(rownames(lacmres3r),lacmres3f[,"0.4"],type="b",col="red")
lines(rownames(lacmres3r),lacmres3f[,"0.3"],type="b",col="blue")
lines(rownames(lacmres3r),lacmres3f[,"0.2"],type="b",col="darkgreen")
grid()

# 100
load("GSE158422_simulate100.Rdata")

gres2[,"PREC"] <- gres2[,"TP"] / ( gres2[,"TP"] + gres2[,"FP"] )
gres3p <- do.call(rbind,lapply(groupsizes, function (g) { gres2[params$groupsizes==g,"PREC"] }))
colnames(gres3p) <- deltas
rownames(gres3p) <- groupsizes
gres3p %>% kbl(caption="GSAmeth precision") %>% kable_paper("hover", full_width = F)

gres3r <- do.call(rbind,lapply(groupsizes, function (g) { gres2[params$groupsizes==g,"REC"] }))
colnames(gres3r) <- deltas
rownames(gres3r) <- groupsizes
gres3r %>% kbl(caption="GSAmeth recall") %>% kable_paper("hover", full_width = F)

gres3f <- F1(gres3p,gres3r)
gres3f %>% kbl(caption="GSAmeth F1") %>% kable_paper("hover", full_width = F)

plot(rownames(gres3p),gres3p[,"0.5"],type="b",las=1,ylab="precision",xlab="",
  main="",ylim=c(0,1))
lines(rownames(gres3p),gres3p[,"0.4"],type="b",col="red")
lines(rownames(gres3p),gres3p[,"0.3"],type="b",col="blue")
lines(rownames(gres3p),gres3p[,"0.2"],type="b",col="darkgreen")
grid()

legend("bottomright", legend=c("0.5", "0.4", "0.3", "0.2"),title="delta",
  lty=1, lwd=2, col=c("black", "red", "blue", "darkgreen"), cex=1)

plot(rownames(gres3r),gres3r[,"0.5"],type="b",las=1,ylab="recall",xlab="group size",
  main="GSA",ylim=c(0,1))
lines(rownames(gres3r),gres3r[,"0.4"],type="b",col="red")
lines(rownames(gres3r),gres3r[,"0.3"],type="b",col="blue")
lines(rownames(gres3r),gres3r[,"0.2"],type="b",col="darkgreen")
grid()

plot(rownames(gres3r),gres3f[,"0.5"],type="b",las=1,ylab="F1",xlab="",
  main="",ylim=c(0,1))
lines(rownames(gres3r),gres3f[,"0.4"],type="b",col="red")
lines(rownames(gres3r),gres3f[,"0.3"],type="b",col="blue")
lines(rownames(gres3r),gres3f[,"0.2"],type="b",col="darkgreen")
grid()

lacmres2[,"PREC"] <- lacmres2[,"TP"] / ( lacmres2[,"TP"] + lacmres2[,"FP"] )
lacmres3p <- do.call(rbind,lapply(groupsizes, function (g) { lacmres2[params$groupsizes==g,"PREC"] }))
colnames(lacmres3p) <- deltas
rownames(lacmres3p) <- groupsizes
lacmres3p %>% kbl(caption="LAM precision") %>% kable_paper("hover", full_width = F)

lacmres3r <- do.call(rbind,lapply(groupsizes, function (g) { lacmres2[params$groupsizes==g,"REC"] }))
colnames(lacmres3r) <- deltas
rownames(lacmres3r) <- groupsizes
lacmres3r %>% kbl(caption="LAM recall") %>% kable_paper("hover", full_width = F)

lacmres3f <- F1(lacmres3p,lacmres3r)
lacmres3f %>% kbl(caption="LAM F1") %>% kable_paper("hover", full_width = F)

plot(rownames(lacmres3p),lacmres3p[,"0.5"],type="b",las=1,ylab="precision",xlab="",
  main="",ylim=c(0,1))
lines(rownames(lacmres3p),lacmres3p[,"0.4"],type="b",col="red")
lines(rownames(lacmres3p),lacmres3p[,"0.3"],type="b",col="blue")
lines(rownames(lacmres3p),lacmres3p[,"0.2"],type="b",col="darkgreen")
grid()

legend("bottomright", legend=c("0.5", "0.4", "0.3", "0.2"),title="delta",
  lty=1, lwd=2, col=c("black", "red", "blue", "darkgreen"), cex=1)

plot(rownames(lacmres3r),lacmres3r[,"0.5"],type="b",las=1,ylab="recall",xlab="group size",
  main="LAM",ylim=c(0,1))
lines(rownames(lacmres3r),lacmres3r[,"0.4"],type="b",col="red")
lines(rownames(lacmres3r),lacmres3r[,"0.3"],type="b",col="blue")
lines(rownames(lacmres3r),lacmres3r[,"0.2"],type="b",,col="darkgreen")
grid()

plot(rownames(lacmres3r),lacmres3f[,"0.5"],type="b",las=1,ylab="F1",xlab="",
  main="",ylim=c(0,1))
lines(rownames(lacmres3r),lacmres3f[,"0.4"],type="b",col="red")
lines(rownames(lacmres3r),lacmres3f[,"0.3"],type="b",col="blue")
lines(rownames(lacmres3r),lacmres3f[,"0.2"],type="b",col="darkgreen")
grid()

# png
png(filename="fig2.png",height=150,width=210,units="mm",res=200)

par(mar = c(4.1, 4.1, 2.1, 0.5))

par(mfrow=c(2,3))

plot(rownames(gres3p),gres3p[,"0.5"],type="b",las=1,ylab="precision",xlab="",
  main="",ylim=c(0,1))
lines(rownames(gres3p),gres3p[,"0.4"],type="b",col="red")
lines(rownames(gres3p),gres3p[,"0.3"],type="b",col="blue")
lines(rownames(gres3p),gres3p[,"0.2"],type="b",col="darkgreen")
grid()

legend("bottomright", legend=c("0.5", "0.4", "0.3", "0.2"),title="delta",
  lty=1, lwd=2, col=c("black", "red", "blue", "darkgreen"), cex=1)

plot(rownames(gres3r),gres3r[,"0.5"],type="b",las=1,ylab="recall",xlab="group size",
  main="GSA",ylim=c(0,1))
lines(rownames(gres3r),gres3r[,"0.4"],type="b",col="red")
lines(rownames(gres3r),gres3r[,"0.3"],type="b",col="blue")
lines(rownames(gres3r),gres3r[,"0.2"],type="b",col="darkgreen")
grid()

plot(rownames(gres3r),gres3f[,"0.5"],type="b",las=1,ylab="F1",xlab="",
  main="",ylim=c(0,1))
lines(rownames(gres3r),gres3f[,"0.4"],type="b",col="red")
lines(rownames(gres3r),gres3f[,"0.3"],type="b",col="blue")
lines(rownames(gres3r),gres3f[,"0.2"],type="b",col="darkgreen")
grid()

plot(rownames(lacmres3p),lacmres3p[,"0.5"],type="b",las=1,ylab="precision",xlab="",
  main="",ylim=c(0,1))
lines(rownames(lacmres3p),lacmres3p[,"0.4"],type="b",col="red")
lines(rownames(lacmres3p),lacmres3p[,"0.3"],type="b",col="blue")
lines(rownames(lacmres3p),lacmres3p[,"0.2"],type="b",col="darkgreen")
grid()

legend("bottomright", legend=c("0.5", "0.4", "0.3", "0.2"),title="delta",
  lty=1, lwd=2, col=c("black", "red", "blue", "darkgreen"), cex=1)

plot(rownames(lacmres3r),lacmres3r[,"0.5"],type="b",las=1,ylab="recall",xlab="group size",
  main="LAM",ylim=c(0,1))
lines(rownames(lacmres3r),lacmres3r[,"0.4"],type="b",col="red")
lines(rownames(lacmres3r),lacmres3r[,"0.3"],type="b",col="blue")
lines(rownames(lacmres3r),lacmres3r[,"0.2"],type="b",,col="darkgreen")
grid()

plot(rownames(lacmres3r),lacmres3f[,"0.5"],type="b",las=1,ylab="F1",xlab="",
  main="",ylim=c(0,1))
lines(rownames(lacmres3r),lacmres3f[,"0.4"],type="b",col="red")
lines(rownames(lacmres3r),lacmres3f[,"0.3"],type="b",col="blue")
lines(rownames(lacmres3r),lacmres3f[,"0.2"],type="b",col="darkgreen")
grid()

dev.off()

par(mfrow=c(1,1))

# pdf
png(filename="fig2.pdf",height=10,width=15)

par(mar = c(4.1, 4.1, 2.1, 0.5))

par(mfrow=c(2,3))

plot(rownames(gres3p),gres3p[,"0.5"],type="b",las=1,ylab="precision",xlab="",
  main="",ylim=c(0,1))
lines(rownames(gres3p),gres3p[,"0.4"],type="b",col="red")
lines(rownames(gres3p),gres3p[,"0.3"],type="b",col="blue")
lines(rownames(gres3p),gres3p[,"0.2"],type="b",col="darkgreen")
grid()

legend("bottomright", legend=c("0.5", "0.4", "0.3", "0.2"),title="delta",
  lty=1, lwd=2, col=c("black", "red", "blue", "darkgreen"), cex=1)

plot(rownames(gres3r),gres3r[,"0.5"],type="b",las=1,ylab="recall",xlab="group size",
  main="GSA",ylim=c(0,1))
lines(rownames(gres3r),gres3r[,"0.4"],type="b",col="red")
lines(rownames(gres3r),gres3r[,"0.3"],type="b",col="blue")
lines(rownames(gres3r),gres3r[,"0.2"],type="b",col="darkgreen")
grid()

plot(rownames(gres3r),gres3f[,"0.5"],type="b",las=1,ylab="F1",xlab="",
  main="",ylim=c(0,1))
lines(rownames(gres3r),gres3f[,"0.4"],type="b",col="red")
lines(rownames(gres3r),gres3f[,"0.3"],type="b",col="blue")
lines(rownames(gres3r),gres3f[,"0.2"],type="b",col="darkgreen")
grid()

plot(rownames(lacmres3p),lacmres3p[,"0.5"],type="b",las=1,ylab="precision",xlab="",
  main="",ylim=c(0,1))
lines(rownames(lacmres3p),lacmres3p[,"0.4"],type="b",col="red")
lines(rownames(lacmres3p),lacmres3p[,"0.3"],type="b",col="blue")
lines(rownames(lacmres3p),lacmres3p[,"0.2"],type="b",col="darkgreen")
grid()

legend("bottomright", legend=c("0.5", "0.4", "0.3", "0.2"),title="delta",
  lty=1, lwd=2, col=c("black", "red", "blue", "darkgreen"), cex=1)

plot(rownames(lacmres3r),lacmres3r[,"0.5"],type="b",las=1,ylab="recall",xlab="group size",
  main="LAM",ylim=c(0,1))
lines(rownames(lacmres3r),lacmres3r[,"0.4"],type="b",col="red")
lines(rownames(lacmres3r),lacmres3r[,"0.3"],type="b",col="blue")
lines(rownames(lacmres3r),lacmres3r[,"0.2"],type="b",,col="darkgreen")
grid()

plot(rownames(lacmres3r),lacmres3f[,"0.5"],type="b",las=1,ylab="F1",xlab="",
  main="",ylim=c(0,1))
lines(rownames(lacmres3r),lacmres3f[,"0.4"],type="b",col="red")
lines(rownames(lacmres3r),lacmres3f[,"0.3"],type="b",col="blue")
lines(rownames(lacmres3r),lacmres3f[,"0.2"],type="b",col="darkgreen")
grid()

dev.off()

```

## Session information

```{r,save}

sessionInfo()

```

## Notes

LA

* simla: parametric self-contained

* simlac: parametric competitive

* simnla: nonparametric self-contained

* simnlac: nonparametric competitive

AL

* simal: parametric self-contained

* simalc: parametric competitive

* simnal: nonparametric self-contained

* simnalc: nonparametric competitive
